Smoothing seismicity techniques applied to seismic source characterization and probabilistic seismic hazard analysis in Brazil

### 2015-02-03 manuscript.tex
fixing typos and cleaning for now

% Glossaries
%\usepackage[sanitize=none,acronym,toc]{glossaries}
\usepackage[acronym,toc]{glossaries}
%\usepackage[acronym,toc]{glossaries}
% Define a new glossary type
\newglossary[slg,toc]{symbols}{sym}{sbl}{Lista de Símbolos}
\newglossary[elg]{equations}{eqn}{eql}{Equations}
\makeglossaries

@@

\authors{Marlon Pirchiner\altaffilmark{1,2}}

@@

Intraplate seismicity is a hard problem to solve in tectonics, mainly during the lack of observed seismicity in many areas where seismic hazard assesment are also difficult and the methods applied are often close correlated to the geometries defined by the experts with valuable criteria.

Three smoothing seismicity techniques are reviewed and their application to the Brazilian seismicity and a seismic hazard assesment are discussed.

@@

This last one was proposed initialy to take care of the background seismicity rate in the context of long-term earthquake forecast and the time-independent assumption about the annual background seismic rate could be used
(not indiscriminately REFERENCE) as a grid of points seismic sources.

@@

All the hazard calculations was performed on OpenQuake open source software.

@@

The Brazilian seismic hazard building code \citep{nbr_15421_2006} was based on the GSHAP \citep{giardini_1999} results for South America \citep{shedlock_tanner_1999}, which is the last published model for the whole region. Howerver, some significant seismic areas (Porto dos Gaúchos \citep{barros_2009}), are not represented in the GPSHAP-based building codes.

@@

To perform the \gls{kde}, both its shape and bandwidth needs to be defined. Most kernels are gaussian or power-laws or one of their variants. The main distinction between the methods described here is the bandwidth selection process.

\citet{frankel_1995} smoothing method applies a gaussian \citet{nadaraya_1964}-\citet{watson_1964} smoothing technique with a \emph{fixed} smoothing-distance bandwidth to estimate the seismicity rate. \citet{woo_1996} proposes a \emph{magnitude-dependent} kernel bandwidth based on the nearest neighbour mean distance in magnitude bins. \citet{helmstetter_2012} proposes a \emph{locally-adaptive} bandwidth, based on kernel estimations for the background seismicity rate in long-term forecasts on both space and time dimensions.

@@

\subsection{Specifical Purpose}

@@

\citet{frankel_1995} smoothing seismicity proposal consisted originally in using a "correlation distance" $d_F$ as the \textit{fixed} kernel bandwidth and applying the Nadaraya-Watson \citep{nadaraya_1964, watson_1964} estimator to smooth a 2D seismicity histogram using a gaussian kernel:
\begin{equation}
	\ensuremath{
		\tilde{n}_j = \frac{ \sum_{i} n_i \,e^{ - \left(\frac{\gls{sym:dij}}{\gls{sym:dF}}\right)^2}}
						   { \sum_{i}     e^{ - \left(\frac{\gls{sym:dij}}{\gls{sym:dF}}\right)^2}},
	}
	\label{eq:ni}
\end{equation}
where $\tilde{n}_j$ is the smoothed seismicity (number of earthquakes with magnitude $m$ above the minimum magnitude \gls{sym:Md} in the catalog) on cell $j$. $n_i$ is the earthquake counting in each other cell $i$ and \gls{sym:dij} is \glsdesc{sym:dij}.

@@

\citet{woo_1996} suggested to evaluate the contribution of each earthquake $i$, located at $\boldsymbol{r}_i$, to the seismicity rate $R$ (number of earthquakes per year and unity area) on the cell centered in $\boldsymbol{r}$, depending on the magnitude $m$:
\begin{equation}
	\ensuremath{
		\gls{sym:Rrm} = \sum_{i=1}^{N} \frac{ K(\gls{sym:r} - \gls{sym:ri}, m)}
											{T({\gls{sym:ri})}},
	}
	\label{eq:Rrm}
\end{equation}
where $N$ is the number of earthquakes $i$ in the magnitude bin $m \pm \mathrm{d}m$,
and $T(\gls{sym:ri})$ is the completess time of magnitude $m$ observed on \gls{sym:ri}.

@@

Just to compare, in figure \ref{fig_woo_bandwidth} it is shown the functions computed by \citet{beauval_2003} for Norway and Spain. The angular coeficient ($a_1$) is in the same range of the others but the general mean distance between events into the same magnitude bin is higher than other countries. It is in some way expected. Brazil is quite large than mentioned countries.

@@

Even work on a forecast perspective, the background seismicity rate for a long-term time-independent forecast could be used to charachterize a diffused seismicity under the common assumption that this background seismic rate is invariant over time.

@@

\subsubsection{Maximum likelyhood optimization}

@@

If the earthquake occurrence could be modelled by a Poisson process with rate $N_p$, then the probability to observe exactly $n$ events on the considered time frame is
\begin{equation}
	\ensuremath{
		\gls{sym:pNn} = \frac{{N_p}^n e^{-N_p}}{n!}.
	}
	\label{eq:poisson_probability}
\end{equation}

@@

Referencing equation. Mauris vel lorem magna, tristique auctor ipsum. Aliquam pharetra eleifend massa. Donec porttitor sagittis luctus. Aliquam pretium luctus leo quis congue. Morbi vel felis mi. Referencing Table \ref{fig_seismicity}. Referencing Figure \ref{placeholder}.

@@

Vivamus magna enim, aliquet id cursus a, pharetra ut purus. Phasellus suscipit nisi iaculis mi vulputate id interdum velit dictum. Nam ullamcorper elit in lectus ultrices vitae volutpat massa gravida. Etiam sagittis commodo neque eget placerat. Sed et nisi faucibus metus interdum adipiscing id nec lacus. Donec ipsum diam, malesuada at euismod consectetur, placerat quis diam. Phasellus cursus semper viverra. Proin magna tortor, blandit in ultricies id, facilisis at nibh. Proin eu neque est. Etiam euismod auctor ante. Mauris mauris sem, tincidunt a placerat rutrum, porta id est. Aenean non velit porta eros condimentum facilisis at in nibh. Etiam cursus purus ut orci rhoncus sit amet semper eros porttitor. Etiam ac leo at ipsum tincidunt consequat ac non sapien. Aenean sed leo diam, venenatis pharetra odio.

\subsubsection{Simulation 2}

Suspendisse viverra eleifend nulla at facilisis. Nullam eget tellus orci. Cras sit amet lorem velit. Maecenas rhoncus pellentesque orci eget vulputate. Phasellus massa nisi, mattis nec elementum accumsan, blandit non neque. In ac enim elit, sit amet luctus ante. Cras feugiat commodo lectus, vitae convallis dui sagittis id. In in tellus lacus, sed lobortis eros. Phasellus sit amet eleifend velit. Duis ornare dapibus porttitor. Maecenas eros velit, dignissim at egestas in, tincidunt lacinia erat. Proin elementum mi vel lectus suscipit fringilla. Mauris justo est, ullamcorper in rutrum interdum, accumsan eget mi. Maecenas ut massa aliquet purus eleifend vehicula in a nisi. Fusce molestie cursus lacinia.

\subsection{Real Data}

Aliquam interdum pellentesque scelerisque. Sed tincidunt suscipit purus, id aliquet nulla vehicula quis. Duis sed nisl lorem. Vivamus erat ante, dignissim et aliquam vel, adipiscing vitae magna. Cras id dapibus metus. Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus. Proin ut lectus ut nisi congue ullamcorper. Ut ac turpis ligula. Sed faucibus bibendum nunc eget gravida.

@@

Nam fermentum sapien at enim varius consectetur. Quisque lobortis imperdiet mauris, et accumsan libero vulputate vitae. Integer lacinia purus vel metus tempus suscipit. Curabitur ac sapien quis mauris euismod commodo. Sed pharetra sem elit. Fusce ultrices, mauris eu fermentum tempor, tellus sem ornare lectus, in convallis nunc urna id dolor. Donec convallis ligula vitae sem viverra fermentum. Mauris in ullamcorper erat. Donec ultrices tempus nibh quis vestibulum.

Praesent volutpat, nibh in dignissim commodo, tellus justo consequat erat, vel consequat mi arcu vel lectus. Aliquam a tellus nec felis sagittis consequat. Quisque convallis imperdiet neque a tempor. Nulla non erat urna. Mauris vel lorem magna, tristique auctor ipsum. Aliquam pharetra eleifend massa. Donec porttitor sagittis luctus. Aliquam pretium luctus leo quis congue. Morbi vel felis mi. Suspendisse viverra tortor pretium orci lacinia eleifend. Phasellus aliquam, nunc eu cursus feugiat, erat odio porttitor libero, quis accumsan orci ipsum ut lorem. Vestibulum pharetra malesuada egestas. Sed non orci sit amet erat suscipit fringilla in et diam. Vestibulum ante ipsum primis in faucibus orci luctus et ultrices posuere cubilia Curae; Nunc ut rhoncus nulla. Aenean porta rhoncus suscipit.

Vivamus magna enim, aliquet id cursus a, pharetra ut purus. Phasellus suscipit nisi iaculis mi vulputate id interdum velit dictum. Nam ullamcorper elit in lectus ultrices vitae volutpat massa gravida. Etiam sagittis commodo neque eget placerat. Sed et nisi faucibus metus interdum adipiscing id nec lacus. Donec ipsum diam, malesuada at euismod consectetur, placerat quis diam. Phasellus cursus semper viverra. Proin magna tortor, blandit in ultricies id, facilisis at nibh. Proin eu neque est. Etiam euismod auctor ante. Mauris mauris sem, tincidunt a placerat rutrum, porta id est. Aenean non velit porta eros condimentum facilisis at in nibh. Etiam cursus purus ut orci rhoncus sit amet semper eros porttitor. Etiam ac leo at ipsum tincidunt consequat ac non sapien. Aenean sed leo diam, venenatis pharetra odio.

Suspendisse viverra eleifend nulla at facilisis. Nullam eget tellus orci. Cras sit amet lorem velit. Maecenas rhoncus pellentesque orci eget vulputate. Phasellus massa nisi, mattis nec elementum accumsan, blandit non neque. In ac enim elit, sit amet luctus ante. Cras feugiat commodo lectus, vitae convallis dui sagittis id. In in tellus lacus, sed lobortis eros. Phasellus sit amet eleifend velit. Duis ornare dapibus porttitor. Maecenas eros velit, dignissim at egestas in, tincidunt lacinia erat. Proin elementum mi vel lectus suscipit fringilla. Mauris justo est, ullamcorper in rutrum interdum, accumsan eget mi. Maecenas ut massa aliquet purus eleifend vehicula in a nisi. Fusce molestie cursus lacinia.

@@

Vivamus magna enim, aliquet id cursus a, pharetra ut purus. Phasellus suscipit nisi iaculis mi vulputate id interdum velit dictum. Nam ullamcorper elit in lectus ultrices vitae volutpat massa gravida. Etiam sagittis commodo neque eget placerat. Sed et nisi faucibus metus interdum adipiscing id nec lacus. Donec ipsum diam, malesuada at euismod consectetur, placerat quis diam. Phasellus cursus semper viverra. Proin magna tortor, blandit in ultricies id, facilisis at nibh. Proin eu neque est. Etiam euismod auctor ante. Mauris mauris sem, tincidunt a placerat rutrum, porta id est. Aenean non velit porta eros condimentum facilisis at in nibh. Etiam cursus purus ut orci rhoncus sit amet semper eros porttitor. Etiam ac leo at ipsum tincidunt consequat ac non sapien. Aenean sed leo diam, venenatis pharetra odio.

>>>>

% Glossaries
%\usepackage[sanitize=none,acronym,toc]{glossaries}
\usepackage[acronym,toc]{glossaries}
%\usepackage[acronym,toc]{glossaries}
% Define a new glossary type
\newglossary[slg,toc]{symbols}{sym}{sbl}{Symbols List}
\newglossary[elg]{equations}{eqn}{eql}{Equations}
\makeglossaries

@@

\authors{Marlon Pirchiner\altaffilmark{1,2} 
		and Marcelo Assumpção\altaffilmark{1}
		}

@@

Intraplate seismicity is a hard problem to solve in tectonics, mainly during the lack of observed seismicity in many areas where seismic hazard assessment are also difficult and the methods applied are often close correlated to the geometries defined by the experts with valuable criteria.

Three smoothing seismicity techniques are reviewed and their application to the Brazilian seismicity and a seismic hazard assessment are discussed.

@@

This last one was proposed initially to take care of the background seismicity rate in the context of long-term earthquake forecast and the time-independent assumption about the annual background seismic rate could be used
(not indiscriminately REFERENCE PAGANI WEATHERILL) as a grid of points seismic sources.

@@

All the hazard calculations was performed on OpenQuake \citep{pagani_2014} open source software.

@@

The Brazilian seismic hazard building code \citep{nbr_15421_2006} was based on the GSHAP \citep{giardini_1999} results for South America \citep{shedlock_tanner_1999}, which is the last published model for the whole region. However, some significant seismic areas (Porto dos Gaúchos \citep{barros_2009}), are not represented in the GPSHAP-based building codes.

@@

To perform the \gls{kde}, both its shape and bandwidth needs to be defined. Most kernels are Gaussian or power-laws or one of their variants. The main distinction between the methods described here is the bandwidth selection process.

\citet{frankel_1995} smoothing method applies a Gaussian \citet{nadaraya_1964}-\citet{watson_1964} smoothing technique with a \emph{fixed} smoothing-distance bandwidth to estimate the seismicity rate. \citet{woo_1996} proposes a \emph{magnitude-dependent} kernel bandwidth based on the nearest neighbor mean distance in magnitude bins. \citet{helmstetter_2012} proposes a \emph{locally-adaptive} bandwidth, based on kernel estimations for the background seismicity rate in long-term forecasts on both space and time dimensions.

@@

\subsection{Specific Purpose}

@@

\citet{frankel_1995} smoothing seismicity proposal consisted originally in using a "correlation distance" $d_F$ as the \textit{fixed} kernel bandwidth and applying the Nadaraya-Watson \citep{nadaraya_1964, watson_1964} estimator to smooth a 2D seismicity histogram using a Gaussian kernel:
\begin{equation}
	\ensuremath{
		\tilde{n}_j = \frac{ \sum_{i} n_i \,e^{ - \left(\frac{\gls{sym:dij}}{\gls{sym:dF}}\right)^2}}
						   { \sum_{i}     e^{ - \left(\frac{\gls{sym:dij}}{\gls{sym:dF}}\right)^2}},
	}
	\label{eq:ni}
\end{equation}
where $\tilde{n}_j$ is the smoothed seismicity (number of earthquakes with magnitude $m$ above the minimum magnitude \gls{sym:Md} in the catalog) on cell $j$. $n_i$ is the earthquake counting in each other cell $i$ and \gls{sym:dij} is \glsdesc{sym:dij}.

@@

\citet{woo_1996} suggested to evaluate the contribution of each earthquake $i$, located at $\boldsymbol{r}_i$, to the seismicity rate $R$ (number of earthquakes per year and unity area) on the cell centered in $\boldsymbol{r}$, depending on the magnitude $m$:
\begin{equation}
	\ensuremath{
		\gls{sym:Rrm} = \sum_{i=1}^{N} \frac{ K(\gls{sym:r} - \gls{sym:ri}, m)}
											{T({\gls{sym:ri})}},
	}
	\label{eq:Rrm}
\end{equation}
where $N$ is the number of earthquakes $i$ in the magnitude bin $m \pm \mathrm{d}m$,
and $T(\gls{sym:ri})$ is the completeness time of magnitude $m$ observed on \gls{sym:ri}.

@@

Just to compare, in figure \ref{fig_woo_bandwidth} it is shown the functions computed by \citet{beauval_2003} for Norway and Spain. The angular coefficient ($a_1$) is in the same range of the others but the general mean distance between events into the same magnitude bin is higher than other countries. It is in some way expected. Brazil is quite large than mentioned countries.

@@

Even work on a forecast perspective, the background seismicity rate for a long-term time-independent forecast could be used to characterize a diffused seismicity under the common assumption that this background seismic rate is invariant over time.

@@

\subsubsection{Maximum likelihood optimization}

@@

If the earthquake occurrence could be modeled by a Poisson process with rate $N_p$, then the probability to observe exactly $n$ events on the considered time frame is
\begin{equation}
	\ensuremath{
		\gls{sym:pNn} = \frac{{N_p}^n e^{-N_p}}{n!}.
	}
	\label{eq:poisson_probability}
\end{equation}

@@

.

@@

.

@@

.

@@

.
